---
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: true
    citation_package: natbib
fontsize: 12pt
geometry: margin=0.75in
bibliography: White_bib.bib
header-includes: 
    \usepackage{float}
    \renewcommand{\thepage}{S\arabic{page}} 
    \renewcommand{\thesection}{S\arabic{section}}  
    \renewcommand{\thetable}{S\arabic{table}}  
    \renewcommand{\thefigure}{S\arabic{figure}}
---

\vspace{2cm}

\begin{center}
 \textbf{Supplementary Material: Article title here}
 
Authors: Easton R. White$^{1*}$
\vspace{3 mm}

Address: \\ \emph{$^1$ Center for Population Biology, University of California-Davis, Davis, California 95616 USA}

*Corresponding author: eawhite@ucdavis.edu

Dec XX, 2016
 \end{center}

\vspace{2cm}

This supplementary material includes:

1. Mathematical description of our model
2. Inverse pattern-orientated estimation procedure


Code for all the figures and tables can be found on [Github](https://github.com/erwhite1).

\vspace{2cm}


## Detailed example of subsampling and power calculations

Here, we illustrate in detail how we performed the subsampling and power calculations for a specific population. As an example, we examine a 35-year time series of (). Simple linear regression indicates a significant decline for this population with a blank and blank. We assume that this significant decline over 35 years is in fact the ``true trend". In statistical jargon, the 35-year trend is an effect that is actually present. We can then use this as a benchmark to see if subsamples of the time series show a similar result.

We first extract all contiguous subsamples of the time series. This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample. We can call each set of subsamples, of the same length, a set. For each subsample, we conduct linear regression and extract model coefficients and p-values. Then, the fraction of subsamples within a set that show significant trends (signfiiant slope coefficient) is the statistical power. It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series.

We can then plot statistical power as a function of time series length. As expected, we can see that power increases with the more years that are sampled. 

Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statisical power of 0.8 implies...

With statistical power of 0.8, we then determine the minimum time series length ($t_min$) required to achieve that level of statistical power. Here, $t_min$ is the first point as which all points to the right are above 0.8. 


```{r, echo=F}
# show an example of subsampling and power calculation routines
```


## Results from other questions of interest


### Growth rate calculations

```{r, echo=F, fig.cap='duh\\label{fig:growth_rate}'}
#Example of using population growth rate instead of power
setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
source("scripts/calculate_geometric_growth.R")
load('cleaned-data/pop_info_example_goodNAs.Rdata')

par(mfrow=c(1,2))
  pop = subset(long_dat,long_dat$ID=='10696') #13192 for Harp seals
  plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,ylim=c(0,2000),las=1)
    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
    mtext(text = 'year',1,line=3,cex=1.4)
    mtext(text = 'population size',2,line=3.5,cex=1.4)

  source('scripts/create_subsamples.R')
  aaa=calculate_average_geometric_growth(pop$popvalue)
  plot(rowMeans(aaa,na.rm=T),cex.axis=1.2,pch=16,ylab='',xlab='',las=1)
    mtext(text = 'years sampled',1,line=3,cex=1.4)
    mtext(text = 'population growth rate',2,line=3.7,cex=1.4)
```




### Using Generalized additive model to identify significant trends

TEXT HERE ON IDEA


```{r,echo=F,fig.cap='results here from GAM.\\label{fig:gam_example}'}
# show gam example side-by-side plot
```

```{r,echo=F,fig.cap='results here from GAM.\\label{fig:gam_result}'}
# show gam results for minimium time required
```


### References
